Standardization of ceramic shape: A case study from the Iron Age pottery from northeastern Taiwan

### 2020-08-21 analysis/paper/paper.Rmd
numerous edits to the text after revisions in response to peer reviews

This research investigates if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, which may indicate social changes in Indigenous societies. We predict that competition within the Indigenous community at Kiwulan for foreign resources and trade partnerships with European or Chinese colonizers may have led to the emergence of craft specialization caused by greater economic and social control of ceramic production by a small group of individuals. Using standardization in ceramic shapes as a proxy for craft specialization, we ask: Did colonial trade impact the shape of locally-produced Indigenous pottery vessels? Did pottery shape become more homogeneous after foreign contacts with European colonizers or Chinese immigrants?

@@

Geometric morphometrics (GMM) differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to quantify and analyze shape [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes. There are two common morphometric methods: landmark and outline approaches [@Adams2004]. Landmark GMM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points. Generalized Procrustes analysis (GPA) is used to superimpose landmark data on a common coordinate system by translating, rotating, and scaling [@Bookstein1991]. After the procedure of GPA, superimposed landmark coordinates become shape variables that allow further statistical analyses [@Slice2007]. A common procedure is using dimensional reduction techniques, such as Principal Components Analysis or Canonical Variate Analysis, to capture the key features that represent the overall shape. Visualization of the reduced data enables the identification of groups, followed by statistical tests to robustly distinguish them. Landmark-based morphometrics have been widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone or metal tools [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with distinct components [@Selden2019; @Topi2017]. GMM is often used to answer research questions related to lithic typological and technological change [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova2018], animal domestication or mobility [@Haruda2019; @Owen2014], or hominid activities through cutmarks and taphonomic traces [@Aramendi2019; @Courtenay2019].

Key questions in archaeological shape analysis normally involve measuring shape standardization over time, or between geographical areas. Standardization is often investigated using multivariate analysis of shape variables computed from landmark data, coupled with coefficients of variation on associated metric data, especially for lithic assemblages. For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to increased maintenance of finished points. @Buchanan2018 analyzed lithic morphology with metric data and identified a more uniform base-shape of Folsom points compared to Clovis points across the western US [@Buchanan2019]. With similar methods, @Smith2016 found standardized bases of fluted point in Alaska and northern Yukon that might indicate a risk management strategy to ensure the ease of replacement during long-distance travel. Other factors, such as low levels of cultural innovation in a small group, could also lead to an increase in standardization of point shapes [@Okumura2014]. To test the effectiveness of measuring standardization, @Birch2019 compared landmark-based GMM to traditional metric analysis with CV using European iron weapons as an example. They demonstrated that landmark-based GMM can capture more variation in not only overall shape, but also bilateral symmetry.

For ceramics, @Topi2017 identified that two types of the Casas Grandes vessels in northwest Mexico tend to have standardized shapes using coefficients of variation for the positions of semi-landmarks across shape groups. They suggested standardization might hint the presence of specialized producers, reflecting social complexity. Another way to explore standardization is pairwise testing of variations in morphological disparity between shape groups by calculating their distances in morphospace, an n-dimensional space where shape groups occupy [@Wills2001]. In this manner, @Selden2018ceramic2 examined Caddo ceramics in northeast Texas using semi-landmark approaches and found am increase in shape standardization over time, providing a basis for further discussion of craft specialization or group identity [@Selden2019]. Similarly, a lithic case study suggests that the Gahagan bifaces from the central Texas exhibit less size standardization than those from the southern Caddo area, indicating different uses or tool types [@Selden2020]. Other applications, such as the study of cranial deformation, demonstrate that landmark approaches with multivariate analyses of shape variances are useful to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007].

A key limitation of landmark approaches in archaeology is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves if not mathematically-defined. In those cases, outline approaches, such as Elliptic Fourier Analysis (EFA), is more effective for assessing morphological variations in the whole structure of two-dimensional closed shapes [@Cardillo2010]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008]. That is, coordinates along a curve are converted into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982]. The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014]. The first systematic use of Fourier series to analyze shapes of artifacts in archaeology was @Gero1984's study of lithic flakes in Peru. They successfully identified the changes in tool shape from a more angular to rounded shape over time. Later, @Saragusti2005 introduced more functions allowing the calculation of the specific shape attributes, such as symmetry, roughness, and deformation. This presents the potential of EFA for the analysis of curves in detail. @Iovictua2009 demonstrated a protocol, including outline digitization, EFA procedure, and multivariate linear regression, to compare resharpening trajectories of European Middle Paleolithic stone tools. They found that resharpening can be independent of morphology, suggesting functional attributes should be studied separately. Recent case studies further support that EFA is applicable and effective to examine lithic assemblages, such as typological classification of Late Woodland points [@Fox2015], the function of flaked obsidian tools in Easter Island [@Lipo2016], the shape and symmetry standardization of the British Acheulean [@Hoggard2019], and the cultural taxonomies for the European Late Palaeolithic [@Ivanovaite2020].

Despite rare ceramic studies using EFA, this approach is promising for analyzing ceramic taxonomy and standardization. For example, @Wilczek2014 evaluated the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. The results present that the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports the claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes. Furthermore, @Wilczek2014 pointed out the potential of EFA for detecting the level of ceramic standardization. In this paper we use EFA to evaluate the level of standardization of ceramics data from Kiwulan, northeastern Taiwan, in relation to the foreign colonial presences to gain insights into the emergence of ceramic specialization. Since our ceramic sample lacks sufficient landmarks that could be identified straightforwardly due to their globular body, EFA is a suitable method by focusing the overall shape of an artifact. In addition, we use a significance test for the equality of coefficients of variation of shape variables to statistically compare the vessel standardization from different periods.

@@

Ceramics analyzed in this study come from 40 units (4m by 4m each) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)). The chronology of the archaeological deposits consists of two cultural components, the upper and the lower, with a sterile layer in between [@Chen2007]. We focus on the upper component, dated from AD 1350 to 1850, spans the late Iron Age and the historical period. The historical period in Taiwan started with the presence of the Europeans in the early 17th century. The Dutch first occupied southern Taiwan in 1624, followed by the Spanish in northern Taiwan in 1626 [@Andrade2007]. In 1642, the Spanish was expelled by the Dutch, who then occupied the previous Spanish forts at Helping Dau in Keelung, and in Tamsui. Western Taiwan remained under Dutch colonial rule until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].

@@

The archaeological record of Kiwulan's upper component shows traces of foreign contact, including Europeans in the 17th century, and waves of Chinese immigrants in the 19th century. Imported ceramics from mainland China, stoneware, and ornaments such as beads have been recovered in the upper component, indicating frequent long-distance trade activities with Europeans and Chinese merchants. Archaeological features such as burials, middens, and post-holes with _in-situ_ posts are widespread across the 1-2 m thick deposit of the upper component, and demonstrate that Kiwulan was a continuously occupied large settlement site [@Chen2007]. To compare different foreign influences, we classified the upper component into three chronological phases: pre-European, European, and Chinese. These phases were identified according to chronologically diagnostic artifacts based on their fined-grained time scale. Despite the availability of 17 radiocarbon ages [@Chen2007], the chronological model demonstrates that time range are too coarse to provide sufficient information for the separation of three phases [@WangandMarwick2020]. Thus, radiocarbon ages only serve as a cross-validation for the diagnostic artifacts if applicable. The diagnostic artifacts include blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles employed by the Chinese in the 19th century [@Chen2007; @Hsieh2009; @Wang2011]. We also examined excavation depth measurement and stratigraphic details reported by the excavators (color, texture, disturbance, etc.) to reliably separate three phases. The deposit exhibited signs of continuous human occupation in each of the three phases with no apparent breaks. More details for the assignment of different phases are in the Online Supplementary Materials.

The most abundant artifacts in the upper component are locally manufactured ceramics, which are distributed throughout the temporal sequence, and across the study area. More than 550,000 sherds were revealed, and around 1,200 vessels could be completely or partially reconstructed (i.e. complete rim or base). There are two shapes of locally-manufactured vessels; a cooking pot and a steamer made of two cooking pots stacked together with a clay filter between. Those vessel shapes demonstrate suites of standard morphological components. Each has a globular body with a short neck and wide mouth (Figure \@ref(fig:KWL-pots)). The exterior surface below the neck is decorated with a variety of impressed geometric motifs. These vessels were potentially used for cooking indicated by evidence of the frequent presence of charred residues and carbon deposits on vessel interiors, and soot on vessel exteriors. Firing resulted in orange and brownish color with a fully oxidized core, or a reduced core with oxidized fringes [@Chen2007]. The vessels were believed to be made with pinching technique according to some hand-shaped traces on vessel interiors, such as finger impressions and seams. This kind of vessel has been widely found at archaeological sites during the late Iron Age and the historical period throughout the Yilan Plain [@NMTH2005].

@@

The sample consists of `r total_no` reconstructed vessels with rim, body and base parts, and we securely provenanced to pre-contact (n = `r pre_no`), post-European (n = `r Euro_no`), and Chinese contact contexts (n = `r Chi_no`).

@@

We used scans of pottery drawings at 300 dpi acquired from the Bureau of Cultural Affairs in Yilan (Figure \@ref(fig:KWL-pots)). All drawings provide a two-dimensional view of vessel cross-sections based on metric measurements. The scanned drawings were imported into Inkscape (http://inkscape.org) for digitization where outlines were manually traced. In those instances where only one side of the cross-section image was available, or some sections were missing, we interpolated the curves and then mirrored and joined to create a closed outline for each vessel. Analyses were conducted in R software (R Core Team, 2019) using functions included in the Momocs package for quantifying and analyzing shapes [@Bonhomme2014]. Outlines were converted into a list of successive x, y pixel coordinates for EFA. We analyzed harmonic coefficients by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation.

@@

As a standardized measure of the spread of data, coefficients of variation (CV) allows a direct comparison for variation in samples measured with different units or means. This is useful to examine the degree of standardization for archaeological assemblages and enables comparison of variation across different sample sizes [@Eerkens2001, pp.498]. Following @Eerkens2001 and @Roux2003, we take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community. Given that CV is robust for positive values due to the representation by a ratio scale, we normalized PC scores to a range between 1 and 10 for the computation of CV.

To answer the question of whether CV values of vessel samples across our three occupational phases are significantly different or not, we used the modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014]. While previous work has used the @Feltz1996's asymptotic test for the equality of coefficients of variation from k populations [@Eerkens2000; @Eerkens2001; @Hoggard2017; @Lycett2008; @Okumura2014], we prefer the MSLR test for shape variables as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].

To complement our investigation of craft specialization through shape standardization, we investigated spatial patterns of ceramic vessels at Kiwulan. As craft specialization increases, pottery distribution we expect a shift from a pattern of vessels dispersed across the site to a pattern of clusters that reflects the loci of production [@Costin2001]. We used a Monte Carlo test for randomness in spatial locations of ceramics to robustly test whether their distribution is significantly clustered or dispersed.

@@

Thirteen harmonics captured 99% of the total harmonic power in the elliptic Fourier coefficients of `r total_no` vessels from three discrete phases. Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot) illustrates differences in vessel shapes described using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods, with the greatest differences evidenced between pre-European and Chinese periods.

@@

```{r, first-two-PCs-byplot-and-tps-plot, echo=FALSE, fig.cap='Left: Significant differences in average vessel shapes between the Chinese and the post-European period using thin plate splines (TPS) with outline deformations required to pass from an extreme of one morphospace to another. Right: Pottery shape distribution by each occupation phase according to the first two PCs.'}
knitr::include_graphics(here("analysis/figures/iso_pca_composit_crop.png"))
```

@@

To compare pottery shape standardization across the three phases we investigated the distributions of the first three PC scores, taking the PC scores as proxy variables for vessel shape (Figure \@ref(fig:PCs-distribution)). The CVs calculated of the three PC support a general trend toward a more standardized shape over time, especially the shape identified by PC1 that represents vessel height and roundness. The PC1 shows a higher variation in the pre-European period and post-European period compared to the Chinese period. That is, a more standardized shape found in the Chinese period. However, the PC2 presents a similar diversity in ceramics assemblages across three phases, while the PC3 demonstrates a slightly standardized shape in the Chinese period.

To see whether the differences in the distribution of PCs between any two phases are substantive or due to chance, we assessed the equality of CVs for PC1 and PC2 with a modified signed-likelihood ratio test  [@Krishnamoorthy2014; @Marwick2019]. P-values for PC1 show significant differences in shape standardization across periods, between Chinese contact with either pre-European or post-European (Table \@ref(tab:CV-test-PCs-table)). This result statistically supports the observation of a higher standardized shape in the Chinese period.

```{r PCs-distribution, fig.cap= "The distribution of normalized PC scores by phases with CV values (%)"}

@@

PC1_PC2_cv_tests <-
bind_rows(PC1_cv_tests, 
          PC2_cv_tests, .id = "PC") %>% 
  mutate(PC = str_glue('PC{PC}')) %>% 
  mutate(PC = as.character(PC)) %>% 
  mutate_at(c("MSLRT", "p_value"), round, 4)

@@

Vessel size is another important variable for detecting standardization, we used the body diameter of vessels as a proxy of size to examine their variations and relationships with vessel shape. We measured body diameter directly from each physical vessel in the collection, and we focus on this metric because it is available for more vessels than any other metric. The body diameter of vessels from the Chinese period is larger than those from the two earlier periods, and the vessels before European contact have the smallest body diameter on average (Figure \@ref(fig:size-shape-regression-plots): A). To investigate vessel form standardization, represented by shape and size, we compared CV for PC1 (shape) in relation to CV for body diameter (size). The result (Figure \@ref(fig:size-shape-regression-plots): B) presents a higher standardization in vessel form in the Chinese period according to smaller CV values compared to those from the other two phases. However, there are no obvious differences in form standardization before and after the European presence. To understand the relationship between shape and size, we modeled linear regression for PCs and the measurement of body diameter, with an examination of Pearson's correlation coefficient and p-value (Figure \@ref(fig:size-shape-regression-plots): C). The results demonstrate that shape changes with size in all phases indicated by moderate positive relationships (0.3 ≤ r ≤ 0.7) and small p-values (≤ 0.05), except for PC1 in the Chinese period and PC2 in the pre-European period. In general, the shorter vessels are larger in body diameter according to the significant positive correlation. However, vessels from the Chinese period do not show this pattern. For the relationship between PC2 and body diameter, the negative relationship suggests that vessels with a narrower neck and mouth tend to have a larger body diameter.

@@

```{r, size-shape-regression-plots, echo=FALSE, fig.cap= "A: Distribution of the body diameter of vessels with coefficients of variation by phases. B: Coefficients of variation of vessel shape represented by PC1 in relation to vessel size represetnted by body diameter, showing a more standardization in vessel form in the Chinese period. C: Correlation between shape variables (PC1 and PC2) and body diameter of vessles with Person's r and P-values by phases"}
knitr::include_graphics(here("analysis/figures/size-boxplot-regression.jpg"))
```

@@

```{r kernel-plot, fig.cap= "A: The spatial distribution of the pottery selected for shape analysis. The quantity is indicated by the color scale. B: Kernel density map visulizes the probability of the density of pottery across space. The maps show a major core area during the pre-European period, multiple core areas during the European period, and a single core during the Chinese period. The bandwidth is based on Silverman (1986)"}

>>>>

This research investigates if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century, or Chinese in the 19th century. These were the two major foreign influences in early historical Taiwan that may relate to social changes in Indigenous societies. We predict that competition within the Indigenous community at Kiwulan for foreign resources and trade partnerships with European or Chinese colonizers may have led to the emergence of craft specialization, caused by greater economic and social control of ceramic production by a small group of individuals. Using standardization in ceramic shapes as a proxy for craft specialization, we ask: Did colonial trade impact the shape of locally-produced Indigenous pottery vessels? Did pottery shape become more homogeneous after foreign contacts with European colonizers or Chinese immigrants?

@@

Geometric morphometrics (GMM) differs from traditional linear measurements through its use of Cartesian coordinates of morphological structures to quantify and analyze shape [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates in terms of their unique point locations with respect to numerical values on coordinate axes. There are two common morphometric approaches: landmark and outline methods [@Adams2004]. Landmark GMM approaches assign a set of landmarks and/or semilandmarks onto objects as reference points. Generalized Procrustes analysis (GPA) is used to superimpose landmark data on a common coordinate system by translating, rotating, and scaling [@Bookstein1991]. After the GPA procedure, superimposed landmark coordinates become shape variables that allow further statistical analyses [@Slice2007]. A common procedure is using dimensional reduction techniques, such as Principal Components Analysis or Canonical Variate Analysis, to capture the key features that represent the overall shape. Visualization of the reduced data enables the identification of groups, followed by statistical tests to robustly distinguish them. Landmark-based morphometrics have been widely applied to archaeological objects with obvious morphological features that provide unambiguous reference points for landmark placement, such as tips and edges of stone or metal tools [@Birch2019; @Lycett2013], visually distinctive bone features [@Haruda2019; @Meloro2015], or ceramic assemblages with distinct components [@Selden2019; @Topi2017]. This approach is often used to answer research questions related to lithic typological and technological change [@Doyon2019; @Eren2015; @Selden2018lithic; @Perez2007; @Presnyakova2018], animal domestication or mobility [@Haruda2019; @Owen2014], or hominid activities through cutmarks and taphonomic traces [@Aramendi2019; @Courtenay2019].

Key questions in archaeological shape analysis normally involve measuring shape standardization over time, or between geographical areas. Standardization is often investigated using multivariate analysis of shape variables computed from landmark data, coupled with coefficients of variation on associated metric data, especially for lithic assemblages. For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to increased maintenance of finished points. @Buchanan2018 analyzed lithic morphology with metric data and identified a more uniform base-shape of Folsom points compared to Clovis points across the western US. With similar methods, @Smith2016 found standardized bases of fluted points in Alaska and northern Yukon that might indicate a risk management strategy to ensure the ease of replacement during long-distance travel. Other factors, such as low levels of cultural innovation in a small group, could also lead to an increase in standardization of point shapes [@Okumura2014]. To test the effectiveness of measuring standardization, @Birch2019 compared landmark-based GMM to traditional metric analysis with CVs using European iron weapons as an example. They demonstrated that landmark-based GMM can capture more variation in not only overall shape, but also bilateral symmetry.

For ceramics, @Topi2017 identified that two types of the Casas Grandes vessels in northwest Mexico tend to have standardized shapes, using coefficients of variation for the positions of semi-landmarks across shape groups. They suggested standardization might hint at the presence of specialized producers, reflecting social complexity. Another way to explore standardization is pairwise testing of variations in morphological disparity between shape groups by calculating their distances in morphospace, an n-dimensional space that shape groups occupy [@Wills2001]. In this manner, Seldon [-@Selden2018ceramic2; -@Selden2019] examined Caddo ceramics in northeast Texas using semi-landmark approaches and found an increase in shape standardization over time, providing a basis for further discussion of craft specialization or group identity. Similarly, the Gahagan bifaces from the central Texas exhibit less size standardization than those from the southern Caddo area, indicating different uses or tool types [@Selden2020]. Other applications, such as studies of cranial deformation, demonstrate that landmark approaches with multivariate analyses of shape variances are useful to evaluate shape standardization [@Kuzminsky2016; @Natahi2019; @Perez2007].

A key limitation of landmark approaches in archaeology is that landmarks may be difficult to reproducibly locate for structures that are mostly or entirely curves, if not mathematically-defined. In those cases, outline approaches, such as Elliptic Fourier Analysis (EFA), are more effective for assessing morphological variations in the whole structure of two-dimensional closed shapes [@Cardillo2010]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions [@Adams2004; @Bonhomme2014; @Claude2008]. That is, coordinates along a curve are converted into Fourier function coefficients, called harmonic coefficients or harmonics [@Kuhl1982]. The number of harmonics determines the quality and precision of the geometric representation of an object. The harmonic power, a cumulative sum of squared harmonic coefficients, provides a robust rule for determining the desired number of harmonics [@Bonhomme2014]. The first systematic use of Fourier series to analyze shapes of artifacts in archaeology was @Gero1984's study of lithic flakes in Peru. They successfully identified the changes in tool shape from a more angular to rounded shape over time. Later, @Saragusti2005 introduced more functions allowing the calculation of the specific shape attributes, such as symmetry, roughness, and deformation. This demonstrated the potential of EFA for the analysis of curves in detail. @Iovictua2009 demonstrated a protocol, including outline digitization, EFA procedure, and multivariate linear regression, to compare resharpening trajectories of European Middle Paleolithic stone tools. He found that resharpening can be independent of morphology, suggesting that functional attributes should be studied separately. Recent case studies further support the effectiveness of EFA for examining lithic assemblages, e.g. for typological classification of Late Woodland points [@Fox2015], analysis of the function of flaked obsidian tools in Easter Island [@Lipo2016], study of the shape and symmetry standardization of the British Acheulean [@Hoggard2019], and investigating cultural taxonomies of the European Late Palaeolithic [@Ivanovaite2020].

Despite few ceramic studies using EFA to date, this approach is promising for analyzing ceramic taxonomy and standardization. For example, @Wilczek2014 evaluated the concordance between EFA and Discrete Cosine Transform (DCT), and a traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. They found that the variation indicated by EFA and DCT matches the traditional ceramic typology, which supports the claim that outline-based approaches can be efficiently used for studying variations in ceramic shapes. @Wilczek2014's findings demonstrate the potential of EFA for detecting variation in ceramic standardization. In this paper we use EFA to evaluate the level of standardization of ceramics data from Kiwulan, northeastern Taiwan around the time of foreign colonial presence to gain insight into the emergence of ceramic specialization. The globular shape of the vessels in our sample means that our specimens lacks visually distinctive landmarks, so EFA is an ideal method because of its focus on the overall shape of an artifact. In addition, we use a novel significance test for the equality of coefficients of variation of shape variables to statistically compare vessel standardization from different periods.

@@

Ceramics analyzed in this study come from 40 units (4m by 4m each) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)). The chronology of the archaeological deposits consists of two cultural components, the upper and the lower, with a sterile layer in between [@Chen2007]. We focus on the upper component, dated from AD 1350 to 1850, because it spans the late Iron Age and the historical period. The historical period in Taiwan started with the presence of the Europeans in the early 17th century. The Dutch first occupied southern Taiwan in 1624, followed by the Spanish in northern Taiwan in 1626 [@Andrade2007]. In 1642, the Spanish were expelled by the Dutch, who then took over the Spanish forts at Helping Dau in Keelung, and in Tamsui. Western Taiwan remained under Dutch colonial rule until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].

@@

The archaeological record of Kiwulan's upper component shows traces of foreign contact, including Europeans in the 17th century, and waves of Chinese immigrants in the 19th century. Imported ceramics from mainland China, stoneware, and ornaments such as beads have been recovered in the upper component, indicating frequent long-distance trade activities with Europeans and Chinese merchants. Archaeological features such as burials, middens, and post-holes with _in-situ_ posts are widespread across the 1-2 m thick deposit of the upper component, and demonstrate that Kiwulan was a continuously occupied large settlement site [@Chen2007]. To compare different foreign influences, we classified the upper component into three chronological phases: pre-European, European, and Chinese. These phases were identified according to chronologically diagnostic artifacts. @Chen2007 reports on 17 radiocarbon ages for Kiwulan, and our Bayesian modeling of these ages shows a consistent result with our artefact-based chronology. However, because the three phases are relatively brief and the number of ages is small, radiocarbon modeling is of limited value to chronology building in this case [more details in @WangandMarwick2020]. The diagnostic artifacts include blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles employed by the Chinese in the 19th century [@Chen2007; @Hsieh2009; @Wang2011]. We also examined excavation depth measurements and stratigraphic details reported by the excavators (color, texture, disturbance, etc.) to reliably separate the three phases. The deposit exhibited signs of continuous human occupation in each of the three phases with no apparent breaks. More details for the assignment of different phases are in the Online Supplementary Materials [@WangandMarwick2020].

The most abundant artifacts in the upper component are locally manufactured ceramics, which are distributed throughout the temporal sequence, and across the study area. More than 550,000 sherds were recovered, and around 1,200 vessels could be completely or partially reconstructed (i.e. complete rim or base). There are two shapes of locally-manufactured vessels; a cooking pot and a steamer made of two cooking pots stacked together with a clay filter between. Those vessel shapes demonstrate suites of standard morphological components. Each has a globular body with a short neck and wide mouth (Figure \@ref(fig:KWL-pots)). The exterior surface below the neck is decorated with a variety of impressed geometric motifs. These vessels were likely used for cooking, as indicated by the frequent presence of charred residues and carbon deposits on vessel interiors, and soot on vessel exteriors. Firing resulted in orange and brownish color with a fully oxidized core, or a reduced core with oxidized fringes [@Chen2007]. The vessels were believed to be made with pinching technique according to some hand-shaped traces on vessel interiors, such as finger impressions and seams. This kind of vessel has been widely found at archaeological sites during the late Iron Age and the historical period throughout the Yilan Plain [@NMTH2005].

@@

The sample consists of `r total_no` reconstructed vessels with rim, body and base parts that were securely provenanced to pre-contact (n = `r pre_no`), post-European (n = `r Euro_no`), and Chinese contact contexts (n = `r Chi_no`).

@@

We used 300 dpi scans of pottery drawings acquired from the Bureau of Cultural Affairs in Yilan (Figure \@ref(fig:KWL-pots)). All drawings provide a two-dimensional view of vessel cross-sections based on metric measurements. The scanned drawings were imported into Inkscape (http://inkscape.org) for digitization where outlines were manually traced. In those instances where only one side of the cross-section image was available, or small sections were missing, we interpolated the curves and then mirrored and joined to create a closed outline for each vessel. Analyses were conducted using R software (R Core Team, 2019) with functions from the Momocs package for quantifying and analyzing shapes [@Bonhomme2014]. Outlines were converted into a list of successive x and y pixel coordinates for EFA. We analyzed harmonic coefficients by principal component analysis (PCA) for dimensionality reduction to illustrate the diversity of the shape data and identify major patterns of variation.

@@

As a standardized measure of the spread of data, coefficients of variation (CV) allows a direct comparison for variation in samples measured with different units or means. This is useful to examine the degree of standardization for archaeological assemblages and enables comparison of variation across different sample sizes [@Eerkens2001, pp.498]. Following @Eerkens2001 and @Roux2003, we take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community. Given that CVs are most informative when computed on either all positive values or all negative values, we normalized PC scores to a range between 1 and 10 for the computation of CV.

To answer the question of whether CV values across our three occupational phases are significantly different or not, we used the modified signed-likelihood ratio (MSLR) test for equality of CVs [@Krishnamoorthy2014]. While previous work has used the @Feltz1996's asymptotic test for the equality of coefficients of variation from k populations [@Eerkens2000; @Eerkens2001; @Hoggard2017; @Lycett2008; @Okumura2014], we prefer the MSLR test for shape variables as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions [@Krishnamoorthy2014].

To complement our investigation of craft specialization through shape standardization, we investigated spatial patterns of ceramic vessels at Kiwulan. As craft specialization increases, we expect a shift from a pattern of vessels dispersed across the site to a pattern of clusters that reflects the loci of production [@Costin2001]. We used a Monte Carlo test for randomness in spatial locations of ceramics to robustly test whether their distribution is significantly clustered or dispersed.

@@

Thirteen harmonics captured 99% of the total harmonic power in the elliptic Fourier coefficients of `r total_no` vessels from three phases. Figure \@ref(fig:first-two-PCs-byplot-and-tps-plot) illustrates differences in vessel shapes using thin-plate spline warping for paired periods, pre- and post-European periods, and post-European and Chinese periods, with the greatest differences evidenced between pre-European and Chinese periods.

@@

```{r, first-two-PCs-byplot-and-tps-plot, echo=FALSE, fig.cap='Left: Significant differences in average vessel shapes between the Chinese and the post-European period are visible using thin plate splines (TPS), with outline deformations required to pass from an extreme of one morphospace to another. Right: Pottery shape distribution by each occupation phase according to the first two PCs.'}
knitr::include_graphics(here("analysis/figures/iso_pca_composit_crop.png"))
```

@@

To compare pottery shape standardization across the three phases we investigated the distributions of the first three PC scores, taking the PC scores as proxy variables for vessel shape (Figure \@ref(fig:PCs-distribution)). The CVs calculated of the three PC support a general trend toward a more standardized shape over time, especially the shape identified by PC1 that represents vessel height and roundness. PC1 shows a higher variation in the pre-European period and post-European period compared to the Chinese period. That is, a more standardized shape found in the Chinese period. However, PC2 presents a similar diversity in ceramics assemblages across three phases, while PC3 demonstrates a slightly standardized shape in the Chinese period.

To see whether the differences in the distribution of PCs between any two phases are substantive or due to chance, we assessed the equality of CVs for PC1 and PC2 with a modified signed-likelihood ratio test  [@Krishnamoorthy2014; @Marwick2019]. P-values for PC1 show significant differences in shape standardization across periods, between Chinese contact with either pre-European or post-European (Table \@ref(tab:CV-test-PCs-table)). This result supports the observation of a more highly standardized shape in the Chinese period.

```{r PCs-distribution, fig.cap= "The distribution of normalized PC scores by phases. CV values (%) are shown in the upper right of each plot."}

@@

PC1_PC2_cv_tests <-
bind_rows(PC1_cv_tests, 
          PC2_cv_tests, .id = "PC") %>% 
  mutate(PC = str_glue('PC{PC}')) %>% 
  mutate(PC = as.character(PC)) %>% 
  mutate_at(c("MSLRT", "p_value"), round, 4) %>% 
  rename(`p-value` = p_value )

@@

Vessel size is another important variable for detecting standardization. We used the body diameter of vessels as a proxy of size to examine their variation and relationships with vessel shape. We measured body diameter directly from each physical vessel in the collection, and we focus on this metric because it is available for more vessels than any other metric. The body diameter of vessels from the Chinese period is larger than those from the two earlier periods, and vessels from before European contact have the smallest body diameter on average (Figure \@ref(fig:size-shape-regression-plots): A). To investigate vessel form standardization, represented by shape and size, we compared CV for PC1 (as a shape variable) in relation to CV for body diameter (as a size variable). The result (Figure \@ref(fig:size-shape-regression-plots): B) shows a higher standardization in vessel form in the Chinese period, with smaller CV values compared to those from the other two phases. However, there are no obvious differences in form standardization before and after the European presence. To understand the relationship between shape and size, we computed linear regression models for PCs and body diameter (Figure \@ref(fig:size-shape-regression-plots): C). The results demonstrate that shape and size are positively correlated in all phases, as indicated by moderate positive relationships (0.3 ≤ r ≤ 0.7) and small p-values (≤ 0.05), except for PC1 in the Chinese period and PC2 in the pre-European period. In general, the shorter vessels are larger in body diameter according to the significant positive correlation. However, vessels from the Chinese period do not show this pattern. For the relationship between PC2 and body diameter, the negative relationship suggests that vessels with a narrower neck and mouth tend to have a larger body diameter.

@@

```{r, size-shape-regression-plots, echo=FALSE, fig.cap= "A: Distribution of the body diameter of vessels with coefficients of variation by phases. B: Coefficients of variation of vessel shape represented by PC1 in relation to vessel size represented by body diameter, showing more standardization (lower CV values) in vessel form and size in the Chinese period. C: Correlation between shape variables (PC1 and PC2) and body diameter of vessels with Pearson's r and p-values by phases"}
knitr::include_graphics(here("analysis/figures/size-boxplot-regression.jpg"))
```

@@

```{r kernel-plot, fig.cap= "A: The spatial distribution of the pottery selected for shape analysis. The quantity is indicated by the color scale. B: Kernel density maps visulize the probability of the density of pottery across space. The maps show a major core area during the pre-European period, multiple core areas during the European period, and a single core during the Chinese period. The bandwidth is based on Silverman (1986)"}
